---
title: "TATA Variants"
author: "Jeremy Primus and Kevin Morey"
date: "2/8/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())           # clear variables
# Load the packages into R
library(knitr)

```

## Introduction

Variants of the pNOS promoter were created by implementing 
single-nucleotide changes to the TATA box of the PNOS promoter.  These changes affect the equilibrium binding constant of the promoter/TATA binding protein interaction.


```{r data input, echo = FALSE}
# Data input

# quantified and normalized in Matlab
relativeLuminescence <- c(1, 0.321591, 0.251972, 0.224204, 0.197042,0.158389, 0.0938343, 0.0848641)
stderror <- c(0.1472, 0.0919, 0.0861, 0.0226, 0.3152, 0.1229, 0.0244, 0.0448)
promoterNames <- c("PNOS Consensus", "PNOS 5-8 complement", "Wildtype PNOS", "PNOS TC8-9GG", "PNOS 4bp deletion", "PNOS 1-4 Complement", "PNOS A4G","PNOS A5C")
sequence <- c('TATATATAC', 'CATATTAAC', 'CATAAATTC', 'CATAAATGG', 'GTTAAATTC','GTATAATTC', 'GTATAATTC', 'CATACATTC')

# Bar plot of relative luminesence
par(mar=c(6,4,1,0))
mids <- barplot(relativeLuminescence, main="PNOS TATA Variants", ylim = c(0, 1.1), col = "blue")
text(x = mids, y = 0, srt = 55, adj = 1, labels = promoterNames, cex = 0.8, xpd = TRUE)
# add error bars
arrows(x0 = mids, y0 = relativeLuminescence - stderror/2, x1 = mids, y1 = relativeLuminescence + stderror/2, code = 3, angle = 90, length = 0.15 )

```

  
## Analysis
Assuming the measured luminescence is proportional to the transcription rate, we can approximate the effect of the nucleotide edits on the equilibrium binding constant, $K_{eq}$
     
## Thermodynamically Modeling Transcription Rates
  Define all energies relative to a ground state:  $\Delta G$  
  The TATA binding protein/promoter interaction has two possible states: unbound, and bound  
  The unbound state is the same energetically for all promoter variants, so we choose this to be the ground state: 0
  
  For the *i*-th variant, the binding interaction energy, measured relative to the unoccupied state, 0 is $\Delta G_{i}$  

  The partition function describes the statistical properties of a system in thermodynamic equilibrium.

  $$ P(b) = \frac{e^{-\Delta G_{i}}}{1+e^{-\Delta G_{i}}} $$
   Assuming transcription rate $TR_i$ is proportional to binding probability
   $$ TR_{i} = \alpha \frac{e^{-\Delta G_{i}}}{1+e^{-\Delta G_{i}}} + \epsilon $$
where $\alpha$ and $\epsilon$ are the theoretical maximum and minumum transcription rates   
Assume measured luminescence to be proportional to $TR_{i}$  
so $$ L_i = A\left [\alpha \frac{e^{-\Delta G_{i}}}{1+e^{-\Delta G_{i}}} + \epsilon \right] = \widetilde{\alpha} \frac{e^{-\Delta G_{i}}}{1+e^{-\Delta G_{i}}} + \widetilde{\epsilon} $$
```{r perform fit, message= FALSE, results=FALSE, echo=FALSE}
library(DEoptim)

# Bar plot of luminesence 
data <- relativeLuminescence

# function to be minimized
#fitVars <- c(a, e, Go, delG1 - delG8)
minFcn <- function(fitVars) {
  f <-(fitVars[1]*exp(-fitVars[3])/(1+exp(-fitVars[3])) + fitVars[2] - data[1])^2 + sum((fitVars[1]*exp(-fitVars[3]-fitVars[4:10])/(1+exp(-fitVars[3]-fitVars[4:10])) + fitVars[2] - data[2:8])^2)
  return(f)
}
lower <- rep(0,10)                              # set bounds for fit algorithm
upper <- c(20,4.18046e-05,20,20,20,20,20,20,20,20)
set.seed(1234)                                  # set seed for reproducibility 
outDEoptim <- DEoptim(minFcn, lower, upper, DEoptim.control(itermax = 500))     # run fit
summary(outDEoptim)
paramFits <- outDEoptim$optim$bestmem
```

```{r assign parameters, echo =FALSE}
# assign fit parameters
alpha <- paramFits[1] 
epsilon <- paramFits[2]  
Go <- paramFits[3] 
delG<- paramFits[4:10]  
x = c(Go, Go+delG)
```

```{r plot data, echo = FALSE}
#bar plot of real data
par(mar=c(6,4,1,0))
par(oma=c(1,0,0,0))
plot(x,relativeLuminescence, xaxt = 'n', xlab = "", ylab = "Relative Luminescence", cex=1.2, main = "Thermodynamic Fit")
mtext(text = c(expression(Delta),"     G"), side = 1, line = 0, outer = TRUE, cex = 1.2)
segments(x, 0.0, x, relativeLuminescence, col = "blue", lwd = 12)
text(x = x, y = 0, srt = 45, adj = 1, labels = promoterNames, cex = 0.8, xpd = TRUE)
#add error bars
arrows(x0 = x, y0 = data - stderror/2, x1 = x, y1 = data + stderror/2, code = 3, angle = 90, length = 0.15 )
#overlay fit
xx <-seq(0, 10, length=1000)
y  <-alpha*(exp(-xx)/(1+exp(-xx)))+epsilon
lines(xx,y)


```


  
    
      
  **Table 1.** Binding energy values ($E/k_{B}T$) obtained from thermodynamic fits.  Values are relative to consensus.
```{r table of results, echo=FALSE}
df <- data.frame(sequence, formatC(c(0, delG), digits = 3), row.names = promoterNames)  # create a table
colnames(df) <- c("Sequence", "$\\Delta G$")
kable(df)
```

